A Multi-Scale Analysis of 27,000 Urban Street Networks

### 2017-12-14 draft/v2/latex/paper.tex
revision

\usepackage{moreverb,url}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
		T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\def\volumeyear{2017}

@@

OpenStreetMap offers an underexplored source of worldwide geospatial data useful to urban researchers. This study uses the OSMnx software to automatically download and analyze 27,009 US street networks from OpenStreetMap at metropolitan, municipal, and neighborhood scales---namely, every US city and town ($N=19655$), census urbanized area ($N=497$), and Zillow-defined neighborhood ($N=6857$). It presents wide-ranging empirical findings on US urban form and street network characteristics, emphasizing measures relevant to graph theory, urban design, and morphology such as structure, connectedness, density, centrality, and resilience. In the past, street network data acquisition and processing has been challenging and ad hoc. This study illustrates the use of OSMnx and OpenStreetMap to consistently conduct street network analysis with extremely large sample sizes, with clearly defined network definitions and extents for reproducibility, and using nonplanar, directed graphs. These street networks and measures data sets have been shared in a public repository for other researchers to use.

@@

This study addresses these limitations by conducting a morphological analysis of urban street networks at multiple scales, with large sample sizes, with clearly defined network definitions and extents for reproducibility, and using nonplanar, directed graphs. In particular, it examines 27,009 urban street networks---represented as primal, nonplanar, weighted multidigraphs with possible self-loops---at multiple overlapping scales across the US, focusing on structure, connectedness, centrality, and resilience. It examines the street networks of every incorporated city and town (population size $N=19655$), census urbanized area ($N=497$), and Zillow-defined neighborhood ($N=6857$) in the US. To do so, it uses OSMnx---a new street network research toolkit \cite{redacted_osmnx:_2017}---to download, construct, and analyze these street networks at metropolitan, municipal, and neighborhood scales. These street networks and measures data sets have been compiled and shared in a public repository for other researchers to repurpose at [redacted].

@@

Brief descriptions of these OSMnx-calculated measures are discussed here, but extended technical definitions and algorithms can be found in e.g. \cite{trudeau_introduction_1994, albert_statistical_2002, dorogovtsev_evolution_2002, brandes_network_2005, costa_characterization_2007, newman_structure_2003, newman_networks:_2010, barthelemy_spatial_2011, cranmer_navigating_2017}. \emph{Average street length}, the mean edge length in the undirected representation of the graph, serves as a linear proxy for block size and indicates how fine-grained or coarse-grained the network is. \emph{Node density} is the number of nodes divided by the area covered by the network. \emph{Intersection density} is the density of the set of nodes with more than one street emanating from them (thus excluding dead-ends). \emph{Edge density} is the sum of all edge lengths divided by area, and the physical \emph{street density} is the sum of all edges in the undirected representation of the graph divided by area. These density measures suggest how fine-grained the network is. Finally, \emph{average circuity} represents the average ratio between each edge length and the straight-line distance between the two nodes it links.

@@

Table \ref{table01} presents summary statistics for the entire data set of 497 urbanized areas. These urbanized areas span a wide range of sizes, from the Delano, CA Urbanized Area's 26 km\textsuperscript{2} to the New York--Newark, NY--NJ--CT Urbanized Area's 8,937 km\textsuperscript{2}---thus, density and count-based measures demonstrate substantial variance. Further, these urbanized areas span a wide spectrum of terrains, development eras and paradigms, and cultures.

@@

\caption{Central tendency and statistical dispersion for selected measures of all US urbanized areas' street networks.}
\label{table01}
\begin{tabular}{llllll}

@@

	measure                                     & mean           & std dev        & min            & median         & max            \\

@@

	Area (km\textsuperscript{2})                                  & 460.657        & 858.125        & 25.685         & 184.898        & 8937.429       \\
	Avg of the avg neighborhood degree          & 2.886          & 0.109          & 2.626          & 2.875          & 3.228          \\
	Avg of the avg weighted neighborhood degree & 0.032          & 0.018          & 0.021          & 0.03           & 0.321          \\
	Avg circuity                                & 1.076          & 0.019          & 1.023          & 1.074          & 1.14           \\
	Avg clustering coefficient                  & 0.042          & 0.009          & 0.015          & 0.042          & 0.071          \\
	Avg weighted clustering coefficient         & 0.002          & 0.001          & \textless0.001 & 0.001          & 0.006          \\
	Intersection count                          & 12582          & 26054          & 751            & 4593           & 307848         \\
	Avg degree centrality                       & 0.001          & 0.001          & \textless0.001 & 0.001          & 0.007          \\
	Edge density (km/km\textsuperscript{2})                       & 13.455         & 2.137          & 7.961          & 13.352         & 21.233         \\
	Avg edge length (m)                         & 158.588        & 17.653         & 117.341        & 157.332        & 223.08         \\
	Total edge length (km)                      & 6353           & 12625          & 427            & 2393           & 1.42E+08       \\
	Proportion of dead-ends                     & 0.213          & 0.055          & 0.077          & 0.207          & 0.416          \\
	Proportion of 3-way intersections           & 0.593          & 0.046          & 0.444          & 0.591          & 0.778          \\
	Proportion of 4-way intersections           & 0.187          & 0.063          & 0.054          & 0.178          & 0.422          \\
	Intersection density (per km\textsuperscript{2})              & 26.469         & 6.256          & 12.469         & 26.029         & 49.423         \\
	Average node degree                         & 5.153          & 0.302          & 4.307          & 5.143          & 6.056          \\
	m                                           & 40890          & 83678          & 2516           & 14955          & 981646         \\
	n                                           & 16032          & 32585          & 874            & 5830           & 373309         \\
	Node density (per km\textsuperscript{2})                      & 33.628         & 7.641          & 17.675         & 33.071         & 61.655         \\
	Max PageRank value                          & 0.001          & 0.001          & \textless0.001 & 0.001          & 0.003          \\
	Min PageRank value                          & \textless0.001 & \textless0.001 & \textless0.001 & \textless0.001 & \textless0.001 \\
	Self-loop proportion                        & 0.008          & 0.008          & \textless0.001 & 0.006          & 0.071          \\
	Street density (km/km\textsuperscript{2})                     & 7.262          & 1.221          & 4.217          & 7.171          & 11.797         \\
	Average street segment length (m)           & 161.331        & 17.765         & 119.573        & 160.288        & 225.92         \\
	Total street length (km)                    & 3480           & 7026           & 222            & 1269           & 79046          \\
	Street segment count                        & 22011          & 45725          & 1281           & 7868           & 533757         \\
	Average streets per node                    & 2.764          & 0.162          & 2.223          & 2.77           & 3.217          \\

@@

The typical urbanized area has 26 intersections per km\textsuperscript{2}. Both the densest and the sparsest are in the Deep South: the sparsest is 12.5 (Gainesville, GA urbanized area) and the densest is 49.4 (New Orleans urbanized area). However, New Orleans is an anomaly in the Deep South. Figure \ref{fig01} depicts the intersection density of each American urbanized area: the highest intersection densities concentrate west of the Mississippi River, while the lowest concentrate in a belt running from Louisiana, through the Carolinas and Appalachians, and into New England. In general, only the largest cities on the east coast (e.g., Boston, New York, Philadelphia, Washington) and Florida escape this trend.

\begin{figure*}
	\includegraphics[width=1\textwidth]{fig01.png}
	\caption{Intersection density per urbanized area (left) and average streets per node per urbanized area (right) from lowest (pale yellow) to highest (dark red), in the contiguous US}
	\label{fig01}
\end{figure*}

\begin{figure*}
	\includegraphics[width=0.7\textwidth]{fig02.png}
	\caption{Distribution of node types in 9 urbanized areas, with number of streets emanating from the node on the $x$-axis and proportion of nodes of this type on the $y$-axis }
	\label{fig02}
\end{figure*}

The distribution of node types (i.e., intersections and dead ends) provides an indicator of network connectedness. The typical urbanized area averages 2.8 streets per intersection: many 3-way intersections, fewer dead-ends, and even fewer 4-way intersections. The gridlike San Angelo, TX urbanized area has the most streets per node (3.2) on average, and (outside of Puerto Rico, which contains the seven lowest urbanized areas) the sprawling, disconnected Lexington Park, MD urbanized area has the fewest (2.2). These fit the trend seen in the spatial distribution across the US in Figure \ref{fig01}: urbanized areas in the Great Plains and Midwest have particularly high numbers of streets per node on average, indicating more gridlike, connected networks. Cities in the southern and western US tend to have fewer streets per node, reflecting more dead-ends and a disconnected network. This finding is discussed in more detail in the upcoming section.

In the typical urbanized area, 18\% of nodes are 4-way intersections, 59\% are 3-way intersections, and 21\% are dead-ends. However, this distribution varies somewhat: examining a small sample of nine urbanized areas, chosen to maximize variance, reveals this in clearer detail. In Figure \ref{fig02}, urban Atlanta and Chattanooga have very high proportions of dead-ends---each over 30\% of all nodes---and very few 4-way intersections, indicating a disconnected street pattern. The urbanized areas of Phoenix, Boston, Detroit, and Chattanooga have particularly high proportions of 3-way intersections, each over 60\%, indicating a prevalence of T-intersections. Conversely, Chicago, New Orleans, Duluth, and Lubbock have high proportions of 4-way intersections, indicating more gridlike connected networks. But what is perhaps most notable about Figure \ref{fig02} is that these nine urbanized areas, despite being chosen to maximize variance, are overwhelmingly similar to each other. Every large American urban agglomeration is characterized by a preponderance of 3-way intersections.

@@

\label{table02}

@@

Due to the substantial variation in urbanized area size, from 25 to 9,000 km\textsuperscript{2}, the preceding analysis covers a wide swath of metropolitan types. To better compare apples-to-apples, Table \ref{table02} focuses on the 30 largest urbanized areas cross-sectionally to examine their metric and topological measures. This provides more consistent spatial scales and extents, while offering a window into the similarities and differences in the built forms of America's largest agglomerations.

@@

Table \ref{table03} presents summary statistics of street network characteristics across the entire data set of 19,655 cities and towns---every incorporated city and town in the US. Following recent work by \citet{barthelemy_modeling_2008} and \citet{strano_urban_2013}, we examine the relationship between the total street length $L$ and the number of nodes $n$. The former proposed a model of cities in which $L$ and $n$ scale nonlinearly as $n^{1/2}$, and the latter recently confirmed this empirically with a small sample of ten European cities. However, their small sample size may limit the generalizability of this finding. To investigate this empirically, we examine the relationship between $L$ and $n$ for 19,655 US cities and towns and instead find a strong \emph{linear} relationship ($r^{2}=0.98$), as depicted in Figure \ref{fig03}. We also find a similar linear relationship at the metropolitan and neighborhood scales.

\begin{table*}
\centering
\caption{Selected summary stats for every incorporated city and town in the US.}
\label{table03}
\begin{tabular}{llllll}
	\toprule
	measure                                     & mean    & std dev & min            & median         & max       \\
	\midrule
	Area (km\textsuperscript{2})                                  & 16.703  & 107.499 & 0.039          & 3.918          & 7434.258  \\
	Avg of the avg neighborhood degree          & 2.94    & 0.297   & 0.4            & 2.953          & 3.735     \\
	Avg of the avg weighted neighborhood degree & 0.033   & 0.141   & \textless0.001 & 0.029          & 9.357     \\
	Avg circuity                                & 1.067   & 0.159   & 1              & 1.055          & 20.452    \\
	Avg clustering coefficient                  & 0.048   & 0.041   & \textless0.001 & 0.04           & 1         \\
	Avg weighted clustering coefficient         & 0.01    & 0.018   & \textless0.001 & 0.005          & 0.524     \\
	Intersection count                          & 324     & 1266    & 0              & 83             & 62996     \\
	Avg degree centrality                       & 0.093   & 0.136   & \textless0.001 & 0.052          & 2.667     \\
	Edge density (km/km\textsuperscript{2})                       & 12.654  & 6.705   & 0.006          & 11.814         & 58.603    \\
	Avg edge length (m)                         & 161.184 & 80.769  & 25.822         & 144.447        & 3036.957  \\
	Total edge length (km)                      & 159.067 & 578.521 & 0.052          & 40.986         & 24728.326 \\
	Proportion of dead-ends                     & 0.192   & 0.093   & \textless0.001 & 0.184          & 1         \\
	Proportion of 3-way intersections           & 0.572   & 0.11    & \textless0.001 & 0.579          & 1         \\
	Proportion of 4-way intersections           & 0.237   & 0.129   & \textless0.001 & 0.217          & 1         \\
	Intersection density (per km\textsuperscript{2})              & 29.363  & 21.607  & \textless0.001 & 24.719         & 259.647   \\
	Average node degree                         & 5.251   & 0.668   & 0.8            & 5.268          & 7.166     \\
	m                                           & 1046    & 3924    & 2              & 275            & 176161    \\
	n                                           & 401     & 1516    & 2              & 103            & 71993     \\
	Node density (per km\textsuperscript{2})                      & 35.449  & 24.409  & 0.047          & 30.718         & 296.74    \\
	Max PageRank value                          & 0.034   & 0.046   & \textless0.001 & 0.021          & 0.87      \\
	Min PageRank value                          & 0.005   & 0.018   & \textless0.001 & 0.002          & 0.5       \\
	Self-loop proportion                        & 0.005   & 0.015   & \textless0.001 & \textless0.001 & 1         \\
	Street density (km/km\textsuperscript{2})                     & 6.528   & 3.435   & 0.003          & 6.109          & 29.302    \\
	Average street segment length (m)           & 162.408 & 81.035  & 25.822         & 145.479        & 3036.957  \\
	Total street length (km)                    & 86.096  & 331.048 & 0.026          & 21.005         & 15348.008 \\
	Street segment count                        & 558     & 2208    & 1              & 140            & 107393    \\
	Average streets per node                    & 2.851   & 0.282   & 1              & 2.852          & 4         \\
	\bottomrule
\end{tabular}
\end{table*}

Previous findings \citep[e.g.,][]{masucci_random_2009} also suggest street segment lengths in an urban network follow a power-law distribution. However, we find that these networks instead generally follow a lognormal distribution. This makes theoretical sense as most street networks are not truly scale-free: for example, a typical street network might comprise very few very long street segments (e.g., 1 km), more medium-length segments (e.g., 250 m), many short segments (e.g., 80 m), but very few very short segments (e.g., 10 m). This theoretical illustration suggests the lognormal distribution that this study generally finds across municipal US street networks. An exception, of course, lies in consistently-sized orthogonal grids filling a city's incorporated spatial extents. Such distributions are extremely peaked around a single value: the linear length of a grid block.

We find that such cities are not uncommon in the US, particularly between the Mississippi River and the Rocky Mountains: the Great Plains states are characterized by a unique street network form that is both orthogonal and reasonably dense. The \emph{former} is partly the result of topography (flat terrain that allows idealized grids) and design history (rapid platting and development during the late nineteenth century) that favor orthogonal grids, as discussed earlier. The \emph{latter} results from the fact that most towns across the Great Plains exhibit minimal suburban sprawl. Thus, municipal boundaries snugly embrace the gridlike street network, without extending to accommodate a vast peripheral belt of twentieth century sprawl, circuity, and \enquote{loops and lollipops} \cite{southworth_streets_1997} that characterizes cities in e.g. California that were settled in the same era but later subjected to substantial suburbanization.

\begin{figure*}
	\includegraphics[width=0.7\textwidth]{fig03.png}
	\caption{The linear relationship between total street length and number of nodes in the street networks of 6,857 US neighborhoods (left) and 19,655 US cities and towns (right).}
	\label{fig03}
\end{figure*}

\begin{figure*}
	\includegraphics[width=0.85\textwidth]{fig04.png}
	\caption{Left: contiguous US states by median of mean streets per node in municipal street networks, from lowest (pale yellow) to highest (dark red). Right: contiguous US states by median of mean street segment length in municipal street networks, from longest (pale yellow) to shortest (dark red).}
	\label{fig04}
\end{figure*}

@@

\label{table04}
\begin{tabular}{p{1in}p{1in}p{1in}p{1in}p{1in}}

@@

	AK    & 1.28                        & 2.43                 & 1.1          & 223.5                     \\
	AL    & 9.7                         & 2.64                 & 1.07         & 190.81                    \\

@@

	AZ    & 12.45                       & 2.77                 & 1.08         & 171.8                     \\

@@

	CT    & 28.05                       & 2.7                  & 1.07         & 165.87                    \\

@@

	DE    & 25.3                        & 2.8                  & 1.06         & 127.8                     \\

@@

	GA    & 15.25                       & 2.78                 & 1.07         & 177.5                     \\
	HI    & 8                           & 2.42                 & 1.07         & 177.93                    \\

@@

	MI    & 20.93                       & 2.9                  & 1.05         & 153.5                     \\

@@

	NH    & 12.22                       & 2.69                 & 1.1          & 175.88                    \\

@@

	NM    & 18.5                        & 2.93                 & 1.05         & 152.02                    \\

>>>>

\usepackage{moreverb}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
		T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\def\volumeyear{2017}

@@

OpenStreetMap offers an underexplored source of worldwide geospatial data useful to urban researchers. This study uses the OSMnx software to automatically download and analyze 27,000 US street networks from OpenStreetMap at metropolitan, municipal, and neighborhood scales---namely, every US city and town ($N=19655$), census urbanized area ($N=497$), and Zillow-defined neighborhood ($N=6857$). It presents wide-ranging empirical findings on US urban form and street network characteristics, emphasizing measures relevant to graph theory, urban design, and morphology such as structure, connectedness, density, centrality, and resilience. In the past, street network data acquisition and processing has been challenging and ad hoc. This study illustrates the use of OSMnx and OpenStreetMap to consistently conduct street network analysis with extremely large sample sizes, with clearly defined network definitions and extents for reproducibility, and using nonplanar, directed graphs. These street networks and measures data sets have been shared in a public repository for other researchers to use.

@@

This study addresses these limitations by conducting a morphological analysis of urban street networks at multiple scales, with large sample sizes, with clearly defined network definitions and extents for reproducibility, and using nonplanar, directed graphs. In particular, it examines 27,000 urban street networks---represented as primal, nonplanar, weighted multidigraphs with possible self-loops---at multiple overlapping scales across the US, focusing on structure, connectedness, centrality, and resilience. It examines the street networks of every incorporated city and town (population size $N=19655$), census urbanized area ($N=497$), and Zillow-defined neighborhood ($N=6857$) in the US. To do so, it uses OSMnx---a new street network research toolkit \cite{redacted_osmnx:_2017}---to download, construct, and analyze these street networks at metropolitan, municipal, and neighborhood scales. These street networks and measures data sets have been compiled and shared in a public repository for other researchers to repurpose at [redacted].

@@

Brief descriptions of these OSMnx-calculated measures are discussed here, but extended technical definitions and algorithms can be found in e.g. \cite{trudeau_introduction_1994, albert_statistical_2002, dorogovtsev_evolution_2002, brandes_network_2005, costa_characterization_2007, newman_structure_2003, newman_networks:_2010, barthelemy_spatial_2011, cranmer_navigating_2017}. \emph{Average street length}, the mean edge length in the undirected representation of the graph, serves as a linear proxy for block size and indicates how fine-grained or coarse-grained the network is. \emph{Node density} is the number of nodes divided by the area covered by the network. \emph{Intersection density} is the density of the set of nodes with more than one street emanating from them (thus excluding dead-ends). \emph{Edge density} is the sum of all edge lengths divided by area, and the physical \emph{street density} is the sum of all edges in the undirected representation of the graph divided by area. These density measures suggest how fine-grained the network is. Finally, \emph{average circuity} represents the average ratio between each edge length and the straight-line distance between the two nodes it links (cf. [redacted]).

@@

Table \ref{tab:measures_urban_areas} presents summary statistics for the entire data set of 497 urbanized areas. These urbanized areas span a wide range of sizes, from the Delano, CA Urbanized Area's 26 km\textsuperscript{2} to the New York--Newark, NY--NJ--CT Urbanized Area's 8,937 km\textsuperscript{2}---thus, density and count-based measures demonstrate substantial variance. Further, these urbanized areas span a wide spectrum of terrains, development eras and paradigms, and cultures.

@@

\caption{Central tendency and statistical dispersion for selected measures of all US urbanized areas' street networks: $\mu$ is the mean, $\sigma$ is the standard deviation, and $D$ is the dispersion index $\frac{\sigma ^ 2}{\mu}$.}
\label{tab:measures_urban_areas}
\begin{tabular}{ l r r r r r r }

@@

	measure                                          & $\mu$          & $\sigma$       & min            & median         & max            & $D$            \\

@@

	Area (km\textsuperscript{2})                     & 460.657        & 858.125        & 25.685         & 184.898        & 8937.429       & 1598.539       \\
	Avg of the avg neighborhood degree               & 2.886          & 0.109          & 2.626          & 2.875          & 3.228          & 0.004          \\
	Avg of the avg weighted neighborhood degree      & 0.032          & 0.018          & 0.021          & 0.030          & 0.321          & 0.011          \\
	Avg circuity                                     & 1.076          & 0.019          & 1.023          & 1.074          & 1.140          & \textless0.001 \\
	Avg clustering coefficient                       & 0.042          & 0.009          & 0.015          & 0.042          & 0.071          & 0.002          \\
	Avg weighted clustering coefficient              & 0.002          & 0.001          & \textless0.001 & 0.001          & 0.006          & \textless0.001 \\
	Intersection count                               & 12582          & 26054          & 751            & 4593           & 307848         & 53949.814      \\
	Avg degree centrality                            & 0.001          & 0.001          & \textless0.001 & 0.001          & 0.007          & 0.001          \\
	Edge density (km/km\textsuperscript{2})          & 13.455         & 2.137          & 7.961          & 13.352         & 21.233         & 0.340          \\
	Avg edge length (m)                              & 158.588        & 17.653         & 117.341        & 157.332        & 223.080        & 1.965          \\
	Total edge length (km)                           & 6353           & 12625          & 427            & 2393           & 1.42e8         & 25089.459      \\
	Proportion of dead-ends                          & 0.213          & 0.055          & 0.077          & 0.207          & 0.416          & 0.014          \\
	Proportion of 3-way intersections                & 0.593          & 0.046          & 0.444          & 0.591          & 0.778          & 0.004          \\
	Proportion of 4-way intersections                & 0.187          & 0.063          & 0.054          & 0.178          & 0.422          & 0.021          \\
	Intersection density (per km\textsuperscript{2}) & 26.469         & 6.256          & 12.469         & 26.029         & 49.423         & 1.478          \\
	Average node degree                              & 5.153          & 0.302          & 4.307          & 5.143          & 6.056          & 0.018          \\
	$m$                                              & 40890          & 83678          & 2516           & 14955          & 981646         & 171238.406     \\
	$n$                                              & 16032          & 32585          & 874            & 5830           & 373309         & 66229.939      \\
	Node density (per km\textsuperscript{2})         & 33.628         & 7.641          & 17.675         & 33.071         & 61.655         & 1.736          \\
	Max PageRank value                               & 0.001          & 0.001          & \textless0.001 & 0.001          & 0.003          & \textless0.001 \\
	Min PageRank value                               & \textless0.001 & \textless0.001 & \textless0.001 & \textless0.001 & \textless0.001 & \textless0.001 \\
	Self-loop proportion                             & 0.008          & 0.008          & \textless0.001 & 0.006          & 0.071          & 0.008          \\
	Street density (km/km\textsuperscript{2})        & 7.262          & 1.221          & 4.217          & 7.171          & 11.797         & 0.205          \\
	Average street segment length (m)                & 161.331        & 17.765         & 119.573        & 160.288        & 225.920        & 1.956          \\
	Total street length (km)                         & 3480           & 7026           & 222            & 1269           & 79046          & 14185.880      \\
	Street segment count                             & 22011          & 45725          & 1281           & 7868           & 533757         & 94987.570      \\
	Average streets per node                         & 2.764          & 0.162          & 2.223          & 2.770          & 3.217          & 0.010          \\

@@

The typical urbanized area has 26 intersections per km\textsuperscript{2}. Both the densest and the sparsest are in the Deep South: the sparsest is 12.5 (Gainesville, GA urbanized area) and the densest is 49.4 (New Orleans urbanized area). However, New Orleans is an anomaly in the Deep South. Figure \ref{fig:fig01} depicts the intersection density of each American urbanized area: the highest intersection densities concentrate west of the Mississippi River, while the lowest concentrate in a belt running from Louisiana, through the Carolinas and Appalachians, and into New England. In general, only the largest cities on the east coast (e.g., Boston, New York, Philadelphia, Washington) and Florida escape this trend.

\begin{figure*}
	\includegraphics[width=1\textwidth]{fig01.png}
	\caption{Intersection density per urbanized area (left) and average streets per node per urbanized area (right) from lowest (pale yellow) to highest (dark red), in the contiguous US}
	\label{fig:fig01}
\end{figure*}

\begin{figure*}
	\includegraphics[width=0.7\textwidth]{fig02.png}
	\caption{Distribution of node types in 9 urbanized areas, with number of streets emanating from the node on the $x$-axis and proportion of nodes of this type on the $y$-axis }
	\label{fig:fig02}
\end{figure*}

The distribution of node types (i.e., intersections and dead ends) provides an indicator of network connectedness. The typical urbanized area averages 2.8 streets per intersection: many 3-way intersections, fewer dead-ends, and even fewer 4-way intersections. The gridlike San Angelo, TX urbanized area has the most streets per node (3.2) on average, and (outside of Puerto Rico, which contains the seven lowest urbanized areas) the sprawling, disconnected Lexington Park, MD urbanized area has the fewest (2.2). These fit the trend seen in the spatial distribution across the US in Figure \ref{fig:fig01}: urbanized areas in the Great Plains and Midwest have particularly high numbers of streets per node on average, indicating more gridlike, connected networks. Cities in the southern and western US tend to have fewer streets per node, reflecting more dead-ends and a disconnected network. This finding is discussed in more detail in the upcoming section.

In the typical urbanized area, 18\% of nodes are 4-way intersections, 59\% are 3-way intersections, and 21\% are dead-ends. However, this distribution varies somewhat: examining a small sample of nine urbanized areas, chosen to maximize variance, reveals this in clearer detail. In Figure \ref{fig:fig02}, urban Atlanta and Chattanooga have very high proportions of dead-ends---each over 30\% of all nodes---and very few 4-way intersections, indicating a disconnected street pattern. The urbanized areas of Phoenix, Boston, Detroit, and Chattanooga have particularly high proportions of 3-way intersections, each over 60\%, indicating a prevalence of T-intersections. Conversely, Chicago, New Orleans, Duluth, and Lubbock have high proportions of 4-way intersections, indicating more gridlike connected networks. But what is perhaps most notable about Figure \ref{fig:fig02} is that these nine urbanized areas, despite being chosen to maximize variance, are overwhelmingly similar to each other. Every large American urban agglomeration is characterized by a preponderance of 3-way intersections.

@@

\label{tab:largest_urbanized_areas}

@@

The dispersion index $D$ in Table \ref{tab:measures_urban_areas} demonstrates the heterogeneity of each indicator across the data set. Several \enquote{families} of indicators can be discerned by their heterogeneity. For instance, counts and totals such as $m$, $n$, and the total street length are extremely heterogeneous. Densities and average distances such as intersection density and the average street segment length exhibit only moderate heterogeneity. Finally, several topological measures such as the average clustering coefficient and PageRank are extremely homogeneous. Due to the substantial variation in urbanized area size, from 25 to 9,000 km\textsuperscript{2}, the preceding analysis covers a wide swath of metropolitan types. To better compare apples-to-apples, Table \ref{tab:largest_urbanized_areas} focuses on the 30 largest urbanized areas cross-sectionally to examine their metric and topological measures. This provides more consistent spatial scales and extents, while offering a window into the similarities and differences in the built forms of America's largest agglomerations.

@@

Table \ref{tab:measures_cities} presents summary statistics of street network characteristics across the entire data set of 19,655 cities and towns---every incorporated city and town in the US. Following recent work by \citet{barthelemy_modeling_2008} and \citet{strano_urban_2013}, we examine the relationship between the total street length $L$ and the number of nodes $n$ across different cities. The former proposed a model of city network evolution in which $L$ and $n$ scale nonlinearly as $n^{1/2}$, and the latter suggested that this relationship applies cross-sectionally, using a small empirical sample of ten European cities. However, the latter's small sample size may limit the generalizability of this finding. To investigate this empirically, we examine the relationship between $L$ and $n$ for 19,655 US cities and towns and instead find a strong \emph{linear} relationship ($r^{2}=0.98$), as depicted in Figure \ref{fig:fig03}. We also find a similar linear relationship at the metropolitan and neighborhood scales.
\caption{Central tendency and statistical dispersion for selected measures of all incorporated cities and towns in the US: $\mu$ is the mean, $\sigma$ is the standard deviation, and $D$ is the dispersion index $\frac{\sigma ^ 2}{\mu}$.}
\label{tab:measures_cities}
\begin{tabular}{ l r r r r r r }
	measure                                          & $\mu$   & $\sigma$& min            & median         & max       & $D$      \\
	Area (km\textsuperscript{2})                     & 16.703  & 107.499 & 0.039          & 3.918          & 7434.258  & 691.860  \\
	Avg of the avg neighborhood degree               & 2.940   & 0.297   & 0.400          & 2.953          & 3.735     & 0.030    \\
	Avg of the avg weighted neighborhood degree      & 0.033   & 0.141   & \textless0.001 & 0.029          & 9.357     & 0.607    \\
	Avg circuity                                     & 1.067   & 0.159   & 1.000          & 1.055          & 20.452    & 0.024    \\
	Avg clustering coefficient                       & 0.048   & 0.041   & \textless0.001 & 0.04           & 1.000     & 0.035    \\
	Avg weighted clustering coefficient              & 0.010   & 0.018   & \textless0.001 & 0.005          & 0.524     & 0.033    \\
	Intersection count                               & 324     & 1266    & 0              & 83             & 62996     & 4951.293 \\
	Avg degree centrality                            & 0.093   & 0.136   & \textless0.001 & 0.052          & 2.667     & 0.199    \\
	Edge density (km/km\textsuperscript{2})          & 12.654  & 6.705   & 0.006          & 11.814         & 58.603    & 3.553    \\
	Avg edge length (m)                              & 161.184 & 80.769  & 25.822         & 144.447        & 3036.957  & 40.473   \\
	Total edge length (km)                           & 159.067 & 578.521 & 0.052          & 40.986         & 24728.326 & 2104.061 \\
	Proportion of dead-ends                          & 0.192   & 0.093   & \textless0.001 & 0.184          & 1.000     & 0.045    \\
	Proportion of 3-way intersections                & 0.572   & 0.11    & \textless0.001 & 0.579          & 1.000     & 0.021    \\
	Proportion of 4-way intersections                & 0.237   & 0.129   & \textless0.001 & 0.217          & 1.000     & 0.070    \\
	Intersection density (per km\textsuperscript{2}) & 29.363  & 21.607  & \textless0.001 & 24.719         & 259.647   & 15.900   \\
	Average node degree                              & 5.251   & 0.668   & 0.800          & 5.268          & 7.166     & 0.085    \\
	$m$                                              & 1046    & 3924    & 2              & 275            & 176161    & 14714.556\\
	$n$                                              & 401     & 1516    & 2              & 103            & 71993     & 5734.363 \\
	Node density (per km\textsuperscript{2})         & 35.449  & 24.409  & 0.047          & 30.718         & 296.740   & 16.807   \\
	Max PageRank value                               & 0.034   & 0.046   & \textless0.001 & 0.021          & 0.870     & 0.062    \\
	Min PageRank value                               & 0.005   & 0.018   & \textless0.001 & 0.002          & 0.500     & 0.060    \\
	Self-loop proportion                             & 0.005   & 0.015   & \textless0.001 & \textless0.001 & 1.000     & 0.042    \\
	Street density (km/km\textsuperscript{2})        & 6.528   & 3.435   & 0.003          & 6.109          & 29.302    & 1.807    \\
	Average street segment length (m)                & 162.408 & 81.035  & 25.822         & 145.479        & 3036.957  & 40.433   \\
	Total street length (km)                         & 86.096  & 331.048 & 0.026          & 21.005         & 15348.008 & 1272.917 \\
	Street segment count                             & 558     & 2208    & 1.000          & 140            & 107393    & 8745.983 \\
	Average streets per node                         & 2.851   & 0.282   & 1.000          & 2.852          & 4.000     & 0.028    \\

Previous findings \citep[e.g.,][]{masucci_random_2009} also suggest street segment lengths in an urban network follow a power-law distribution. However, we find that these networks instead generally follow lognormal-style right-skewed distributions. This makes theoretical sense as most street networks are not truly scale-free: for example, a typical street network might comprise very few very long street segments (e.g., 1 km), more medium-length segments (e.g., 250 m), many short segments (e.g., 80 m), but very few very short segments (e.g., 10 m). To test this in a statistically rigorous way, we fit a set of candidate distributions to the street segment lengths of each US city/town. The candidate distributions comprise the lognormal, Gumbel, gamma, exponentiated Weibull, Fréchet, power-law, uniform, and exponential distributions. We then assess each fit with a Kolmogorov-Smirnov test, which tells us how closely each fitted distribution matches the observed data. Power-law distributions provide the best fit for fewer than 0.7\% of these cities. In contrast, the exponentiated Weibull distribution provides the best fit 56\% of the time, followed by the Gumbel (19\%), gamma (11\%), and lognormal (7\%) distributions.

An exception to this general pattern, of course, lies in consistently-sized orthogonal grids filling a city's incorporated spatial extents. Such distributions are extremely peaked around a single value: the linear length of a grid block. We find that such cities are not uncommon in the US, particularly between the Mississippi River and the Rocky Mountains: the Great Plains states are characterized by a unique street network form that is both orthogonal and reasonably dense. The \emph{former} is partly the result of topography (flat terrain that allows idealized grids) and design history (rapid platting and development during the late nineteenth century) that favor orthogonal grids, as discussed earlier. The \emph{latter} results from the fact that most towns across the Great Plains exhibit minimal suburban sprawl. Thus, municipal boundaries snugly embrace the gridlike street network, without extending to accommodate a vast peripheral belt of twentieth century sprawl, circuity, and \enquote{loops and lollipops} \cite{southworth_streets_1997} that characterizes cities in e.g. California that were settled in the same era but later subjected to substantial suburbanization.

\begin{figure*}
	\includegraphics[width=\textwidth]{fig03.png}
	\caption{The linear relationship between total street length and number of nodes in the street networks of 6,857 US neighborhoods (left) and 19,655 US cities and towns (right).}
	\label{fig:fig03}
\end{figure*}

\begin{figure*}
	\includegraphics[width=\textwidth]{fig04.png}
	\caption{Left: contiguous US states by median of mean streets per node in municipal street networks, from lowest (pale yellow) to highest (dark red). Right: contiguous US states by median of mean street segment length in municipal street networks, from longest (pale yellow) to shortest (dark red).}
	\label{fig:fig04}
\end{figure*}

@@

\label{tab:medians_state_aggregate}
\begin{tabular}{ l r r r r }

@@

	AK    &  1.28                       & 2.43                 & 1.10         & 223.50                    \\
	AL    &  9.70                       & 2.64                 & 1.07         & 190.81                    \\

@@

	AZ    & 12.45                       & 2.77                 & 1.08         & 171.80                    \\

@@

	CT    & 28.05                       & 2.70                 & 1.07         & 165.87                    \\

@@

	DE    & 25.30                       & 2.80                 & 1.06         & 127.80                    \\

@@

	GA    & 15.25                       & 2.78                 & 1.07         & 177.50                    \\
	HI    &  8.00                       & 2.42                 & 1.07         & 177.93                    \\

@@

	MI    & 20.93                       & 2.90                 & 1.05         & 153.50                    \\

@@

	NH    & 12.22                       & 2.69                 & 1.10         & 175.88                    \\

@@

	NM    & 18.50                       & 2.93                 & 1.05         & 152.02                    \\
